The unequal distribution of access to education and transport:

### 2016-02-09 paper.Rmd

---
title: 'School accessibility: a new index applied to enrollment and performance inequalities
  in Sao Paulo'
author: "Ana Moreno and Robin Lovelace"
output: pdf_document
bibliography: school-access-bib.bib
---

@@

Travel to school is an everyday reality for millions of young people around the world.
The mode, duration, safety, comfort and pollution levels of this trip has huge
impacts on the future generation, yet has received relatively little academic
attention.
Access to high quality quality schools is highly variable across geographical
space, making travel options are vital for ensuring a more equitable supply of
educational opportunity to diverse groups. Conversely, poor accessibility
to deprived area can reinforce social inequalities, with long-term implications.
Based on this emerging context, this paper develops a school accessibility index
for local areas.
This index is based on current travel patterns, local travel infrastructure serving the school
catchment and the provision of travel services such as public transport links.
Using data from the Sao Paulo Household Travel Survey (2007 and 2012), School Census
and electronic card data, we demonstrate how this index can be used
measure current provision and evaluate future interventions.
The results suggest that that school accessibility can
impact the social background of school attendees and
academic performance.
We conclude that extending the concept of local accessibility indicators
to the 'school commute' can ensure that transport decision makers take
equality and health impacts on young people into account.

@@

$$
\tau=2
$$

@@

@@

>>>>

---
title: 'The unequal distribution of access to education and transport:
  insights from for São Paulo'
author: Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Frederico R. Ramos (Fundação
  Getulio Vargas) and Robin Lovelace (University of Leeds)
output: 
  pdf_document: 
    keep_tex: yes
bibliography: school-access-bib.bib
---

@@

In many large Latin American cities such as São Paulo, growing social
and economic inequalities are embedded through unfair education and
transport systems. Good schools are mostly concentrated in wealthy
areas, while transport links to school are deficient in deprived areas,
exacerbating the issue. Inequalities in educational and transport
infrastructure are mutually reinforcing: the right to mobility is
intrinsically linked to the right to education. This is manifested by
the overlap between recent protests against unwanted changes to public
education and the social movements contesting increases in public
transport fares. Another manifestation is to be found in the concept of
school accessibility. This paper sheds light on the transport-education
inequality nexus with reference to a new school accessibility measure
applied São Paulo. By capturing both the unequal distribution of schools
and transport services across space, the index allows embedded
inequalities to be better understood and, with political will,
contested. Our index combines information on the spatial distribution of
children and adolescents by income categories, the location of existing
schools, the travel patterns of students, and the travel infrastructure
serving the school catchment area into a single measure. The index is
used to measure school accessibility locally across São Paulo, using
data sources from Population and School Censuses, commonly available in
Latin American cities. The results illustrate how existing inequalities
are amplified by variable accessibility to schools across income groups
and geographical space. We conclude that extending the concept of local
accessibility indicators to education can help to both contest and
constructively tackle embedded social inequalities.

@@

We calculate three different accessibility measures for different modes: cumulative, gravitiy-based, and competitive.

The cumulative-opportunity measure for mode $M$ (Boisjoly and El-Geneidy, 2016) is defined as:

$$ CO_{i}^M= \sum_{j=1}^n(S_{j}f(C_{ij}) $$

$$ f(C_{ij}) = \left\{ 
                \begin{array}{ll}
                  1 if C_{ij}<=t\\
                  0 if C_{ij}<t
                \end{array}
              \right.
  $$

This measure counts the number of schools available from one area within a certain travel time by mode $M$ threshold. $C_{ij}$ is the travel cost (measured in time) between the centroid of zone i and the centroid of zone j.

Gravity-based accessibility measure

In spite of its simplicity, Boisjoly and El-Geneidy (2016) have found that the cumulative-opportunity measure (using a 45 minute threshold) is highly correlated with a gravity-based accessibility measure ([@handy1997measuring], [@hansen1959accessibility]), defined as:

$$ A_{i}^M= \sum_{j=1}^n(S_{j}e^{\beta*C_{ij}}) $$

Where $A_{i}^M$ is the gravity accessibility at the centroid of area $i$ to all schools at area $j$ using mode $M$, and $\beta$ is a negative exponential cost function. The cost function, based on a negative exponential decay curve, includes reported number of school trips in the 2007 Origin-Destination Survey. (Here we can also use information from the school census on number of students per school to know how many students the area attracts). The proposed accessibility measure captures the fact that more proximate schools (S) are more attractive than those located further away.

Competitive accessibility measure

The measures proposed above take into account the sptail distribution of schooling opportunities, but not the local demand for schooling. This is particularly relevant for our equity analysis, since it could be the case that higher areas are disproportionally served with respect to the number of potentail students living within a certain travel distance, whereas the opposite holds for lower-income areas.  In order to assess the mismatch between the demand and supply for schooling, we use the sum of students in schools in each area (the demand), and the sum of individuals within the school grade age-group living in each area (the supply) in the following competitive accessibility measure, first proposed by [@shen1998location] and adapted by El-Geneidy et al (2015):

$$ CA_{i}^M= \sum_{j=1}^n\frac{P_{j}-f(C_{ij})}{\sum_{k=1}^n(Y_{k}-f(C_{kj}))}$$

Where the numerator discounts the number of pupils in area $j$ $P_{j}$ by how far area $i$ is from area $j$ using the same function as before, and the denominator discounts the number of students (young adults) living in zone $k$ $Y_{k}$ by how far they are from area $j$. 
In this way, the discounted number of study places at each area is divided by the potential students available to fill those places, and then summed in order to obtain a single accessibility measure for each area $i$.

In order to assess equality in access to schooling, we stratify our population of students in two student groups (we could do it for more income groups): the students living with a head of household who earns 1 minimum wage or less, and those living with a head of household who earns more than 1 minimum wage. The accessibility measures are based on student's travel times by different modes using data from the 2007 Origin-Destination survey. Alternatively, for the case of public transport, wet use actual transit times in 2015 for comparison (bilhete unico data?).

@@

- Descriptive statistics about modal choice and travel time by AEPs and level of income
- Spatial Distribution of Schools and Transit-Dependent Students at the AEP Level

@@

<!--

  -->
